A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-03-24 masters_thesis/si_paper/paper.tex
Fixed a problem with the number of processors we said we used on one function
A lot of work on the paper.  It is very close.
This incorporates a lot of new experiments, plus a lot of changes from
reviewer feedback.

% Better references, I think
\renewcommand{\sec}[1]{Section~\ref{sec:#1}}
\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
\newcommand{\alg}[1]{Algorithm~\ref{alg:#1}}

@@

\institute{M. Gardner \at
Brigham Young University.
3342 TMCB, Provo, UT 84604
Tel.: 801-422-8717
Fax: 801-422-0169
\email{mjg82@byu.edu}
\and
A. McNabb \at
Brigham Young University.
3342 TMCB, Provo, UT 84604
\email{a@cs.byu.edu}
\and
K. Seppi \at
Brigham Young University.
3330 TMCB, Provo, UT 84604
\email{k@cs.byu.edu}
}

@@

In our approach, we refactor PSO such that the computation needed for iteration
$t+1$ can be done in parallel with the computation needed for iteration $t$.
Thus we can perform two iterations of PSO at once.  Even with some amount of
wasted computation, we show that this approach to parallelization in PSO often
outperforms the na\"ive parallelization of simply adding particles to the
swarm.  Our methods produce results that are exactly equivalent to PSO; this is
not a new algorithm or variant, only a new method of parallelization.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{iters_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of iterations of the algorithm performed.}
  \label{fig:iters-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of function evaluations performed.  Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO,
we can introduce new speculative techniques that often out-perform na\"ive
parallelizations of PSO.  These relaxations make better use of the information
obtained from the extra exploration made by the speculative function
evaluations.  We also explore the idea that, like branch prediction in
processors, we need not speculatively evaluate \emph{all} possible future
positions, we can accelerate the algorithm even if we are just \emph{likely} to
have guessed right.  By pruning the speculation to just paths that are
statistically likely to reproduce the paths that are equivalent to PSO we can
increase the swarm size without increasing the number of speculative
evaluations.  We also consider several recovery strategies for cases where the
pruned set of speculative evaluations does not contain the evaluation that
standard PSO would have done.  A further improvement we explore is speculating
several iterations ahead instead of just one, which is made possible by pruning
the number of speculative evaluations.

@@

An attempt to standardize PSO found that although a Complete swarm of 50
particles converged to the global optimum more quickly for many benchmark
functions than the same swarm size with a Ring topology, it was also more
likely to permaturely converge to local optima for other functions.  The study
found no significant improvement for any other swarm size between 20 and 100
and concluded with a recommendation to use a swarm of 50 particles with a Ring
topology as a starting point.  The authors acknowledged that choosing the ideal
topology requires thorough experimentation for the particular
problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
explained that a large swarm with a sparse topology propogates information
slowly~\citep{montes-de-oca-2009-frankensteins-pso}.

@@

There have been several parallel implementations of PSO presented in the
literature.  The improvements described in the literature come in two major
areas: innovations in implementation details and innovations in the use of
topology and swarm size to scale PSO to many processors.

@@

To perform two iterations at a time we must first refactor PSO such that the
determination of the value of the objective function is separate from the rest
of the computation.  For simplicity, this discussion will describe the case
where PSO is performing function minimization using the Ring topology.  In this
example, each particle has two neighbors, the ``right neighbor'' and ``left
neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
respectively.  Though we will only describe the case of the Ring topology, this
method can easily be extended to arbitrary topologies.

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results in \sec{results} show that in many cases it requires
too many extra evaluations to produce competitive results.  On most functions,
the larger swarm size that can be used in na\"ive parallelizations leads to
better performance than is found by trying to do speculative evaluation to
exactly reproduce PSO.  However, if we keep the idea of speculative evaluation
while relaxing the requirement of exactly reproducing the behavior of the
original PSO algorithm, we see some impressive results.

We outline three main improvements to speculative evaluation.  First, in
\sec{pickbest} we describe a method that uses all of the information found in
doing speculative evaluations.  Then Sections~\ref{sec:pruning}
through~\ref{sec:wrong} present a technique that reduces the number of
speculative evaluations that need to be done for each particle.  Finally,
\sec{manyiters} shows a method for speculating several iterations ahead,
instead of just one.

@@

At this point it is also interesting to note a parallel between our methods and
parallel evolution
strategies~\citep{rudolph-1991-distributed-evolution-strategies}.  In evolution
strategies, a parent individual (representing a potential solution to some
objective function) produces a number of offspring by a mutation operator.  One
of the individuals is selected by a selection operator, and that individual
becomes the parent for the next
generation~\citep{beyer-2002-evolution-strategies}.  Our methods are similar,
where our mutation operator is simply the PSO motion equations and the
selection operator is either the indicator function introduced in \sec{sepso},
in the case of our original speculative algorithm, or the standard selector
operator based on fitness, in the case of this Pick Best technique.

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2n+1$.  With a fixed number of processors, this would allow us to greatly
increase the swarm size relative to that needed in the original speculative
algorithm (e.g., with 120 processors, a na\"ive parallelization has a swarm of
size 120, complete speculative evaluation has a swarm of size 15, and pruning
the evaluations to only one per particle allows a swarm of size 60).  As not
all of the branches are evaluated in any given iteration, we call this
technique pruning.

A na\"ive approach to pruning is to keep track of the last branch taken by each
particle and speculate on that branch.  This turns out to be correct far less
than half of the time on average.

@@

We ran experiments to compare our speculative parallelization of PSO to the
na\"ive parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The speculative version actually performs
two iterations of PSO at each ``iteration,'' so we instead call each
``iteration'' a ``time step.''

@@

We also do not report absolute time used by each algorithm, as specific time
values are heavily implementation-dependent.  Using time steps, where at every
time step each processor evaluates the function once, is an abstraction from
individual implementations that still conveys all of the information necessary
to compare algorithms.  In this paper we focus on the effects of algorithm
choices on parallel PSO performance.  As there are inexpensive functions with
complex landscapes and expensive functions with simple landscapes, the behavior
of PSO with respect to the function is the main issue.  For this purpose,
benchmark functions are a useful and efficient tool for understanding the
effects of PSO with expensive objective functions even though the benchmark
functions are themselves inexpensive.  For example, a plot of performance with
respect to iterations for a smooth unimodal function with five minutes per
evaluation would be similar to that of Sphere.

Using the same number of processors for each approach requires that our
speculative parallelizations have a smaller swarm size than the na\"ive
parallelization.  For each of the topologies we used, a particle has three
neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
$7$ speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.  The
extra evaluations required in our speculative approach would instead be used to
evaluate particles in na\"ive parallelizations, so we compare swarms of size
$p$ in speculative evaluation with swarms of size $8p$ in na\"ive approaches.

Where the Complete topology would normally be preferred, we use a Random
topology in our speculative approach, as Complete leads to an explosion in the
number of speculative evaluations.  If speculative evaluation were not being
performed, it is possible that the Complete topology would be used.  However,
the Complete topology also requires a very large amount of interprocessor
communication in distributed PSO, so it is still quite possible that Random
would be used even with na\"ive parallelizations.  But, to be fair in our
comparisons, we compare to the na\"ive parallelization using both the Random
topology and the Complete topology (labeled PSO Random and PSO Complete in our
results).

@@

The $c$ in the function definitions is a shifted center, in order to avoid
origin-seeking bias in the PSO algorithm and its
variants~\citep{monson-2005-origin-seeking-bias}.  We move the center of the
shifted functions (all except Bohachevsky) to be halfway between the center and
the boundary of the initialization region.  For example, if the initialization
region is $[-50,50]^D$, the center is $(25)^D$.  For all of these functions we
used 20 dimensions.

The parallelization techniques we compare are the na\"ive parallelization (here
labeled Na\"ive simply for ease of reference), our original speculative
approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
\sec{relax}.  In presenting our results, we call the approach developed in
\sec{pickbest} Pick Best.  The methods described in \sec{pruning} through
\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
method in \sec{manyiters} we call Many Iterations.

When performing pruning in Pick Best Pruned and Social Promotion Pruned, there
are a large number of ways to prune speculative evaluations.  We experimented
with several, but present results for only one possible pruning.  The pruning
we present uses only the two branches where the $\nbest$ value was not updated:
\casepn\ and \casexn.  Those branches are convenient in that no messages are
needed from neighbors in order to produce the positions of the speculative
particles---in distributed frameworks using several rounds of communication
(see \sec{distributed}), one of the rounds of communication can be dropped
entirely.

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size
$\frac{1}{3}p$ to compare to a swarm of size $p$ with the na\"ive
parallelization, instead of $\frac{1}{8}p$ with other techniques.

There are also many ways to speculate several iterations ahead, and with Many
Iterations we again only show results for one of them.  The combination of
branches we tried uses seven speculative evaluations per particle, matching the
swarm size of the original speculative algorithm.  The seven evaluations we
used corresponded to several iterations of branches \casepn\ and \casexn.  Two
of the evaluations were just one iteration ahead, four were two iterations
ahead, and one was three iterations ahead.  The evaluations that were one
iteration ahead were branches \casepn\ and \casexn; those that went two
iterations ahead were formed by taking either branch \casepn\ or \casexn and
then branch \casepn\ or \casexn; and the evaluation going three iterations
ahead followed branch \casepn\ on all three iterations.

@@

\subsection{Rastrigin}

First we look at Rastrigin.  Rastrigin is a multi-modal function that is best
optimized with a large, Complete swarm.  It has been shown that with Rastrigin,
the more particles there are in the swarm, the lower function value it finds,
up to at least 4000 particles~\citep{mcnabb-2009-large-particle-swarms}.
Smaller swarms get caught in local optima.  Because our speculative algorithms
require significantly smaller swarm sizes, we would expect to not perform very
well on functions such as Rastrigin.  Our experiments show our intuition to be
correct.  In this experiment we used 240 processors, so SEPSO and Pick Best
each had 30 particles, and the Na\"ive algorithms had 240 particles.  As
expected, SEPSO and Pick Best converge quickly to worse local optima than
Na\"ive does.  \fig{rastrigin} shows the results graphically.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin}
  \caption{Function Rastrigin with a swarm that uses 240 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.  All speculative algorithms use the Random topology.}
  \label{fig:rastrigin}
\end{figure}

\subsection{Sphere}

Next we go to Sphere, the simplest of common benchmark functions.  The function
has no local optima (only a single global optimum) and is best optimized in
terms of function evaluations with a small swarm using a Complete topology.  We
expect our methods to be perfect for such functions, and our results show this
intuition again to be correct.  First we compare the basic SEPSO approach to
Na\"ive.  In \fig{basic-sphere} we can see that when both SEPSO and Na\"ive are
using a Random topology, SEPSO clearly outperforms Na\"ive.  However, the
comparison between SEPSO and Na\"ive Complete is not as encouraging---Na\"ive
Complete performs slightly better.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere}
  \caption{Function Sphere with a swarm that uses 240 processors per time step.
  Recall that at each time step all processors evaluate the function once.
  SEPSO uses the Random topology.}
  \label{fig:basic-sphere}
\end{figure}

When we introduce Pick Best, we see more impressive results.  As can be seen in
\fig{sphere-pickbest}, where SEPSO did not manage to perform as well as Na\"ive
Complete, Pick Best handily outperforms it with the same number of processors.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere2}
  \caption{Function Sphere with a swarm that uses 240 processors per time step,
  comparing Pick Best with previous results.  All speculative algorithms use
  the Random topology.}
  \label{fig:sphere-pickbest}
\end{figure}

Still more impressive with Sphere are the results of pruning and speculating
several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
Iterations by far outperforms every other algorithm on this function.  Many
Iterations on average finds values that are more than 40 orders of magnitude
better than Na\"ive using the same number of processors.  This is shown in
\fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
SEPSO, and we do not clutter the graph with a line for it.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere3}
  \caption{Function Sphere with a swarm that uses 240 processors per time step,
  comparing speculating several iterations ahead with previous results.  All
  speculative algorithms use the Random topology.}
  \label{fig:sphere-manyiters}
\end{figure}

\subsection{Schwefel 2.21}

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail to perform as well as Na\"ive because of their smaller swarm size.  In
comparing SEPSO to Na\"ive in \fig{schwefel1}, we can see that SEPSO fails to
perform any better than Na\"ive with a Random topology, and it is by far
outperformed by Na\"ive with a Complete topology.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{schwefel1}
  \caption{Function Schwefel with a swarm that uses 240 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.  SEPSO uses a Random topology.}
  \label{fig:schwefel1}
\end{figure}

As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
perform worse than Na\"ive Complete, with Many Iterations performing comparably
to Na\"ive.  However, one of the speculative algorithms did perform better than
Na\"ive: Pick Best Pruned.  It seems that the larger swarm size available
because of pruning was the cause of the improved performance.  Note that Many
Iterations was initially performing comparably with Pick Best Pruned, but with
more iterations gradually degraded in performance to the level of Na\"ive
Complete.  It is likely that Many Iterations got caught in local optima, and
with more processors would have maintained better performance.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{schwefel2}
  \caption{Function Schwefel with a swarm that uses 240 processors per time
  step, comparing all speculative algorithms.  All speculative algorithms use
  the Random topology.}
  \label{fig:schwefel2}
\end{figure}

\subsection{Griewank}

Next we look at the function Griewank.  It is generally recommended to use the
Ring topology when optimizing the Griewank function, as Complete is prone to
premature convergence on a local
optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  Griewank has a
global optimum with a value of 0, and sometimes the swarm finds the optimum and
sometimes it does not.  Instead of showing average function value at each
iteration, a more informative plot for Griewank shows the percent of runs that
have found the global optimum by each iteration.  We ran 50 trials of each
experiment with Griewank, so that the curves are more smooth.  We show results
in \fig{basic-griewank1} for swarms of size 30 and 240 using the Ring topology.
One can see in the figure that the SEPSO approach finds the optimum faster than
the Na\"ive approach.  However, because the swarm size is so small, SEPSO gets
stuck a little less than half of the time.

>>>>

% Better references, I think
\renewcommand{\sec}[1]{Section~\ref{sec:#1}}
\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
\newcommand{\tabref}[1]{Table~\ref{tab:#1}}
\newcommand{\alg}[1]{Algorithm~\ref{alg:#1}}

@@

\institute{M. Gardner \at
Brigham Young University.
3361 TMCB, Provo, UT 84604
Tel.: 801-422-8717
Fax: 801-422-0169
\email{mjg82@byu.edu}
\and
A. McNabb \at
Brigham Young University.
3361 TMCB, Provo, UT 84604
\email{a@cs.byu.edu}
\and
K. Seppi \at
Brigham Young University.
3361 TMCB, Provo, UT 84604
\email{k@cs.byu.edu}
}

@@

In our approach, we refactor PSO such that the computation needed for iteration
$t+1$ can be done in parallel with the computation needed for iteration $t$.
Thus we can perform two iterations of PSO at once.  Even with some amount of
wasted computation, we show that this approach to parallelization in PSO often
outperforms the naive parallelization of simply adding particles to the
swarm.  Our methods produce results that are exactly equivalent to PSO; this is
not a new algorithm or variant, only a new method of parallelization.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{iters_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of iterations of the algorithm performed.}
  \label{fig:iters-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of function evaluations performed.  Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO,
we can introduce new speculative techniques that often out-perform naive
parallelizations of PSO.  These relaxations make better use of the information
obtained from the extra exploration made by the speculative function
evaluations.  We also explore the idea that, like branch prediction in
processors, we need not speculatively evaluate \emph{all} possible future
positions, we can accelerate the algorithm even if we are just \emph{likely} to
have guessed right.  By pruning the speculation to just paths that are
statistically likely to reproduce the paths that are equivalent to PSO we can
increase the swarm size without increasing the number of speculative
evaluations.  We also consider several recovery strategies for cases where the
pruned set of speculative evaluations does not contain the evaluation that
standard PSO would have done.  A further improvement we explore is speculating
several iterations ahead instead of just one, which is made possible by pruning
the number of speculative evaluations.

@@

An attempt to standardize PSO found that although a Complete swarm of 50
particles converged to the global optimum more quickly for many benchmark
functions than the same swarm size with a Ring topology, it was also more
likely to prematurely converge to local optima for other functions.  The study
found no significant improvement for any other swarm size between 20 and 100
and concluded with a recommendation to use a swarm of 50 particles with a Ring
topology as a starting point.  The authors acknowledged that choosing the ideal
topology requires thorough experimentation for the particular
problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
explained that a large swarm with a sparse topology propagates information
slowly~\citep{montes-de-oca-2009-frankensteins-pso}.

@@

The idea of speculative decomposition in the parallelization literature is not
new~\citep{grama-2003-intro-to-parallel-computing}.  In the field of function
optimization, simulated annealing has previously been parallelized using this
technique~\citep{witte-1991-parallel-simulated-annealing-speculative}, though
we are not aware of other evolutionary or swarm-intelligence based algorithms
having been parallelized with speculative decomposition.

There have been several parallelizations of PSO presented in the literature.
The improvements described in these papers come in two major areas: innovations
in implementation details and innovations in the use of topology and swarm size
to scale PSO to many processors.

@@

To perform two iterations at a time we must first refactor PSO such that the
determination of the value of the objective function is separate from the rest
of the computation.  For simplicity, this discussion will describe the case
where PSO is performing function minimization using the Ring topology.  In this
example, each particle has two neighbors, the ``right neighbor'' and ``left
neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
respectively.  Though we will only describe the case of the Ring topology here,
the math is straightforward for other topologies.  Our discussion of the
implementation in \sec{implementation} is independent of specific topologies,
and we use several different topologies in our experiments.

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results in \sec{results} show that in many cases it requires
too many extra evaluations to produce competitive results.  On most functions,
the larger swarm size that can be used in naive parallelizations leads to
better performance than is found by trying to do speculative evaluation to
exactly reproduce PSO.  However, if we keep the idea of speculative evaluation
while relaxing the requirement of exactly reproducing the behavior of the
original PSO algorithm, we see some impressive results.

We outline three main improvements to speculative evaluation.  First, in
\sec{pickbest} we describe a method that uses all of the information found in
doing speculative evaluations.  Then Sections~\ref{sec:pruning}
through~\ref{sec:wrong} present a technique that reduces the number of
speculative evaluations that need to be done for each particle, allowing
speculative evaluation to use larger swarm sizes with the same number of
processors.  Finally, \sec{manyiters} shows a method for speculating several
iterations ahead, instead of just one.

@@

At this point it is also interesting to note a parallel between our methods and
parallel evolution
strategies~\citep{rudolph-1991-distributed-evolution-strategies}.  In evolution
strategies, a parent individual (representing a potential solution to some
objective function) produces a number of offspring by a mutation operator.  One
of the individuals is selected by a selection operator, and that individual
becomes the parent for the next
generation~\citep{beyer-2002-evolution-strategies}.  Our methods are similar,
where our mutation operator is simply the PSO motion equations and the
selection operator is either the indicator function introduced in \sec{sepso},
in the case of our original speculative algorithm, or the standard selection
operator based on fitness, in the case of this Pick Best technique.

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2n+1$.  With a fixed number of processors, this would allow us to greatly
increase the swarm size relative to that needed in the original speculative
algorithm (e.g., with 120 processors, a naive parallelization has a swarm of
size 120, complete speculative evaluation has a swarm of size 15, and pruning
the evaluations to only one per particle allows a swarm of size 60).  As not
all of the branches are evaluated in any given iteration, we call this
technique pruning.

A naive approach to pruning is to keep track of the last branch taken by each
particle and speculate on that branch.  This turns out to be correct far less
than half of the time on average.

@@

We ran experiments to compare our speculative parallelization of PSO to the
naive parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The speculative version actually performs two
iterations of PSO at each ``iteration,'' so we instead call each ``iteration''
a ``time step.''  As noted in the introduction, this is a reasonable
architecture to use if the function being optimized takes longer than around
100 milliseconds to run.

@@

Using the same number of processors for each approach requires that our
speculative parallelizations have a smaller swarm size than the naive
parallelization.  For each of the topologies we used, a particle has three
neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
$7$ speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.  The
extra evaluations required in our speculative approach would instead be used to
evaluate particles in naive parallelizations, so we compare swarms of size
$p$ in speculative evaluation with swarms of size $8p$ in naive approaches.

\subsection{Running Time}

We do not report absolute time used by each algorithm, as specific time values
are heavily implementation-dependent.  Using time steps, where at every time
step each processor evaluates the function once, is an abstraction from
individual implementations that still conveys all of the information necessary
to compare algorithms.  In this paper we focus on the effects of algorithm
choices on parallel PSO performance.  As there are inexpensive functions with
complex landscapes and expensive functions with simple landscapes, the behavior
of PSO with respect to the function is the main issue.

For this purpose, benchmark functions are a useful and efficient tool for
understanding the effects of PSO with expensive objective functions even though
the benchmark functions are themselves inexpensive.  While benchmark functions
themselves take fractions of a second to evaluate and thus have no need of the
kind of parallelization we explore in this paper, they also have no need of
particle swarm optimization at all, as they can be solved analytically.  Yet
they are useful to us and to optimization researchers generally because they
stand as surrogates for the kinds of functions practitioners are actually
interested in, and they allow us to explore the behavior of optimization
algorithms on several different classes of functions in an easy and
standardized way.

If experimentation with benchmark functions provides sufficient experimental
evidence for algorithmic improvements in serial PSO, it is also sufficient
evidence for algorithmic improvements in parallel PSO.  While the discussion of
exact running times of function evaluations and communication overhead is
certainly appropriate, even essential, to a discussion of specific parallel
implementations, the concern of this paper is more fundamental than
implementation decisions.  No one expects to see actual running times in place
of function evaluations in a paper describing an innovation in the motion
equations of PSO.  Similarly, actual running times should not be required of
those describing algorithmic improvements in parallel PSO.  Just as in a serial
PSO paper, actual running times cloud the more important issues of
implementation-independent properties of the innovations described.

That being said, given the time required for the evaluation of the objective
function and the communication overhead per iteration for a specific
implementation, the actual running time can be recovered from all of our
results.  Simply multiply the number of time steps by the sum of the function
evaluation time and the overhead.  We wish to stress that the ``time steps'' we
report are \emph{exactly} proportional to wall-clock time, given the
architecture we have assumed.

\subsection{Functions Used}

@@

The $c$ in the function definitions is a shifted center, in order to avoid
origin-seeking bias in the PSO algorithm and its
variants~\citep{monson-2005-origin-seeking-bias}.  We move the center of the
shifted functions (all except Bohachevsky) to be halfway between the center and
the boundary of the initialization region.  For example, if the initialization
region is $[-50,50]^D$, the center is $(25)^D$.  We tested all of these
functions in their 20 dimension, 50 dimension, and 500 dimension varieties.

We also tested the parallelization methods on a real problem, that of fitting a
model to a large quantity of data.  We generated 10,000 data points from a
radial basis function network with 10 bases and some added Gaussian noise.  We
then fit a radial basis function network to the data using PSO.  This amounted
to a 30 dimensional function to optimize, with a function evaluation time on
the order of two seconds.

\subsection{Parallelization Techniques Compared}

The parallelization techniques we compare are the naive parallelization (here
labeled Naive simply for ease of reference), our original speculative
approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
\sec{relax}.  In presenting our results, we call the approach developed in
\sec{pickbest} Pick Best.  The methods described in \sec{pruning} through
\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
method in \sec{manyiters} we call Many Iterations.

When performing pruning in Pick Best Pruned and Social Promotion Pruned, there
are a large number of ways to prune speculative evaluations.  We experimented
with several, but present results for only one possible pruning.  The pruning
we present uses only the two branches where the $\nbest$ value was not updated:
\casepn\ and \casexn.  Those branches are convenient in that no messages are
needed from neighbors in order to produce the positions of the speculative
particles---in distributed frameworks using several rounds of communication
(see \sec{distributed}), one of the rounds of communication can be dropped
entirely.  Pruning all but these branches also allows the use of arbitrarily
dense topologies, as the number of speculative particles is no longer dependent
on the number of neighbors the particle has.

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size
$\frac{1}{3}p$ to compare to a swarm of size $p$ with the naive
parallelization, instead of $\frac{1}{8}p$ with other techniques.

There are also many ways to speculate several iterations ahead, and with Many
Iterations we again only show results for one of them.  The combination of
branches we tried uses seven speculative evaluations per particle, matching the
swarm size of the original speculative algorithm.  The seven evaluations we
used corresponded to several iterations of branches \casepn\ and \casexn.  Two
of the evaluations were just one iteration ahead, four were two iterations
ahead, and one was three iterations ahead.  The evaluations that were one
iteration ahead were branches \casepn\ and \casexn; those that went two
iterations ahead were formed by taking either branch \casepn\ or \casexn and
then branch \casepn\ or \casexn; and the evaluation going three iterations
ahead followed branch \casepn\ on all three iterations.  As with Pick Best
Pruned and Social Promotion Pruned, this choice of branches allows the use of
arbitrarily dense topologies, as the number of speculative evaluations per
particle is independent of the number of neighbors the particle has.

\subsection{Topologies}

For each benchmark function we report results using the topology that is widely
considered best for that function, as reported in the literature
(e.g.,~\cite{bratton-2007-defining-a-standard-for-pso}).  In this paper we
limit ourselves to the Ring topology and the Complete topology, as is common
practice, along with the parallel approximation to the Complete topology
mentioned in \sec{pso}.  We also mentioned in \sec{related} that some related
work can be described as changes in topology, particularly that of having
subswarms of fully connected topologies that occasionally communicate with
each other.  This related work focused mainly on functions with very fast
evaluation times where such techniques drastically reduce interprocessor
communication.  With long function evaluations these topologies are not as
practical, as only one particle is evaluated by each processor at each
iteration.  We experimented with a subswarm topology and found that in every
instance except one it performed worse than either Complete or Ring, and thus
we leave it out of the results except in the one instance where it improved
performance.

Where the Complete topology would normally be preferred, we use a Random
topology in SEPSO and Pick Best, as Complete leads to an explosion in the
number of speculative evaluations (as noted, the other speculative techniques
can still use a Complete topology; we often report results on both topologies
for those methods).  If speculative evaluation were not being performed, it is
possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with naive parallelizations~\cite{mcnabb-2009-large-particle-swarms}.  But,
to be fair in our comparisons, we compare to the naive parallelization using
both the Random topology and the Complete topology (labeled PSO Random and PSO
Complete in our results).  Again, the amount of communication overhead is
heavily dependent on implementation details which are not the focus of this
paper.  A practitioner using a particular implementation can compare results
for whichever topology is most practical given the specific implementation.

@@

We first present results in \sec{20dims} for the 20 dimensional variants of the
benchmark functions we tested.  We give a discussion of each function
individually at 20 dimensions, as there are interesting characteristics of the
algorithms that are worth discussing.  In \sec{50dims} we then give a summary
of results for the 50 dimensional variants of the benchmark functions; we do
not go into as much detail in our discussion as the results are very similar to
those in 20 dimensions.  We finish our discussion of benchmark functions with
their 500 dimensional variants in \sec{500dims}, and we discuss results on the
model fitting problem in \sec{model}.

\subsection{20 Dimensions}
\label{sec:20dims}

\subsubsection{Rastrigin}

Rastrigin is a multi-modal function that is best optimized with a large,
Complete swarm.  It has been shown that with Rastrigin, the more particles
there are in the swarm, the lower function value it finds, up to at least 4000
particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
in local optima.  Because our speculative algorithms require significantly
smaller swarm sizes, we would expect to not perform very well on functions such
as Rastrigin.  Our experiments show our intuition to be correct.  In this
experiment we used 240 processors, so SEPSO and Pick Best each had 30
particles, and the Naive algorithms had 240 particles.  As expected, SEPSO and
Pick Best converge quickly to worse local optima than Naive does.
\fig{rastrigin} shows the results graphically, and \tabref{rastrigin-20} shows
results for all of the methods we tried.

\begin{table}
  \caption{Summary of results for function Rastrigin with 20 dimensions,
  measuring number of time steps to reach a value of 20}
  \label{tab:rastrigin-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Naive Ring&25\%&743.4&212.0\\
  \hline
  Naive Random&100\%&372.1&86.6\\
  \hline
  \textbf{Naive Complete}&\textbf{95\%}&\textbf{273.1}&\textbf{99.7}\\
  \hline
  SEPSO Ring&5\%&471.0&0.0\\
  \hline
  SEPSO Random&15\%&208.0&17.0\\
  \hline
  Pick Best Ring&35\%&489.0&152.0\\
  \hline
  Pick Best Random&15\%&136.0&40.1\\
  \hline
  Pick Best Pruned Ring&95\%&545.9&208.5\\
  \hline
  Pick Best Pruned Random&10\%&95.0&11.0\\
  \hline
  Pick Best Pruned Complete&5\%&59.0&0.0\\
  \hline
  Social Promotion Pruned Ring&5\%&580.0&0.0\\
  \hline
  Social Promotion Pruned Random&80\%&278.2&136.3\\
  \hline
  Social Promotion Pruned Complete&50\%&200.5&130.2\\
  \hline
  Many Iterations Ring&10\%&117.0&25.0\\
  \hline
  Many Iterations Random&5\%&57.0&0.0\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rastrigin}
  \caption{Function Rastrigin with a swarm that uses 240 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.  All speculative algorithms use the Random topology.}
  \label{fig:rastrigin}
\end{figure}

\subsubsection{Griewank}
\label{sec:griewank-20}

It is generally recommended to use the Ring topology when optimizing the
Griewank function, as Complete is prone to premature convergence on a local
optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
of Griewank has a value of 0.  When most trials reach the global optimum but a
few get stuck, the resultant ``average value'' graph has a flat line that is
misleading.  Thus instead we present plots showing the percent of trials that
have passed some threshold at each time step.  The threshold we chose for this
case was $10^{-6}$, as that value is below any local optima and the swarm
always successfully reaches 0 once is passes that point.

We ran 50 trials of each experiment with Griewank, so that the curves are more
smooth.  We show results in \fig{basic-griewank1} for swarms of size 30 and 240
using the Ring topology.  One can see in the figure that the SEPSO approach
finds the optimum faster than the Naive approach.  However, because the swarm
size is so small, SEPSO gets stuck a little less than half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.  Instead of showing average function value, we show the percentage of
  runs that have found the global optimum by each iteration.  All algorithms
  use the Ring topology.}
  \label{fig:basic-griewank1}
\end{figure}

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum on average
twice as fast as the Naive approach without a significant loss of accuracy.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per time
  step.  All algorithms use the Ring topology.}
  \label{fig:basic-griewank2}
\end{figure}
